Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-02-21 paper/paper.tex
wrote mannik/stability and triangle sections, added all the relevent data for these sections into the figures of the paper, added all the V shape box data in figures to paper and arrow data to data directory Tested everything on node25.

\section{Stability Analysis and Mannik's Experimental Shapes}
In general we have found that the protein reaction sequence discussed
above results in extremely robust oscillatory behavoir of polar
selection and oscillation in a wide variety of cell shapes.  However
we have found that this robustness is heavily dependent upon cell
size.  Two pole, regular oscillatory behavoir is found in cells which
are long enough in one dimension for the reaction chain to be unstable
but short enough in the perpendicular dimension to be stable.  The
limits are seen when both dimensions are too short to be unstable, so
that initial assymetry in protein concentration relaxes into a
motionless steady state, or when both are too great, so that the
proteins exhibit much more complex and varied behavoir.

In order to explain this theoretically, we have performed stability
analysis of the five differential equations in an infinite slab that
shows a characteristic half wavelength in perturbative spatial
oscillations at which the system becomes unstable.  This length is
dependent upon slab thickness as shown in
Figure~\ref{fig:stability-analysis}.  When increasing the lengths and
widths of our simulated flattened cells, the steady, two pole
oscillations seem to break down when the shortest distance across the
cell is approximately equal to the characteristic half wavelength at
the particular thickness at which we simulate (our simulations of
cells with thickness $0.25\mu m$ corresponds to half wavelengths of
$2.13\mu m$).  Huang~\cite{huang2003dynamic} performs a similar
analysis for a cylindrical shape \fixme{is it cylindrical or 1d?} and
finds characteristic half wavelengths of $2\mu m$.

\begin{figure}
  \includegraphics[width=2.8cm]{../mannik-2.png}
  \includegraphics[width=2.8cm]{../mannik-1.png}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1300-2000-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1100-1100-9500-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1800-2500-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1800-1800-9500-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-2300-3300-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-2800-2800-9500-1500.pdf}
  \caption{Arrow plots for our simulation and mannik's experimental}
  \label{arrow-compare-mannik}
\end{figure}

Figure~\ref{arrow-compare-mannik} shows plots of total MinD
concentration maxima that are global in space and local in time.  The
tip of an arrow represents one of these maxima and the adjacent arrow
(whose tail is touching the previous arrow's tip) points to the next
such maxima in time.  Mannik \emph{et al.} have published similar
plots of two of their cells, and we have attempted to replicate the
shape of these cells in our simulation in order to compare.  We find
that when we replicate the size as well as shape of their cells as
near as possible, we do not see the disordered protein behavoir that
they observe.  Instead we see a robust ability of the cell to find two
prefered pole locations and to oscillate maxima back and forth very
steadily between them.  However, when we keep the same shape but
increase the size, we see that there is a critical size at which the
protein dynamics lose their order and become very complex (showing
hints of prefered wall maxima locations), similar to what Mannik sees
experimentally.  We find that this critical size matches up well with
our stability analysis, in that the critical sizes correspond to sizes
for which the shorter of the two distances across the cell is roughly
equal to the analysis' half wavelength found for this cell thickness.

\begin{figure}
  \includegraphics[width=4.2cm]{../data/shape-randst/plots/box-plot_D--randst-25-800-800-9500-1500}
  \includegraphics[width=4.2cm]{../data/shape-randst/plots/box-plot_D--randst-25-1000-1700-9400-1500}
  \caption{Box plots of Mannik shapes for sizes at the lower stability
    threshhold.  The vertical axis shows stacked and changing over
    time the total number of proteins that are of four different
    compound stages and in the different sections of the cell.  Both
    shapes have longer axis of roughly 2.2$\mu m$.}
  \label{box-mannik}
\end{figure}
Decreasing the cell size in succesive steps, we can see as well the
critical size at which the cell oscillations are quickly damped out.
Figure~\ref{box-mannik} shows box plots of the largest cells simulated
whose oscilltions relax into the steady state solution.  Both have a
long axis of roughly 2.2$\mu m$, in agreement with theory.

@@

Still waiting for the triangle data to show what we expect.  We have
small triangles that gradually go from no oscillations to
oscillations, but the large triangle data does not yet show
irregularities.  I have very large simulations running that should
definitely show this at some point (Give it a few days).

\begin{figure}
  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
  \caption{Box plots of triangle shapes for sizes at the lower stability
    threshhold.  The vertical axis shows stacked and changing over
    time the total number of proteins that are of four different
    compound stages and in the different sections of the cell.}
  \label{box-triangle}
\end{figure}

%% \begin{figure}
%%   %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-400-400-400-1500}
%%   %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-500-300-500-1500}
%%   %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-600-480-360-1500}
%%   \caption{Total protein fluctuation in the right, middle, and left
%%     parts of an equilateral (top), iscoscolese (middle), and 3-4-5
%%     (bottom) triangle.  The vertical axis shows stacked the total
%%     number of proteins that are of four different compound stages and
%%     in the different sections of the cell.}
%%   \label{box-triangle}
%% \end{figure}

@@

\section{Our 'V' Shaped Cells}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9900-1500}
  %%\includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9600-1500}arrow-plot-NflD-randst-25-600-800-9800-1500.pdf

  \caption{Total protein fluctuation in the A shaped shape.  The vertical axis shows stacked the total number
    of proteins that are of four different compound stages and in the
    different sections of the cell.}
  \label{total-oscillation-randst-96-plot}
\end{figure}

@@

\subsubsection{Randst-97-}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-800-600-9700-1500}
  %%\includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9600-1500}
  \caption{Total protein fluctuation in the sideways flying saucer
    pancake cell shape.  The vertical axis shows stacked the total
    number of proteins that are of four different compound stages and
    in the different sections of the cell.}
  \label{box-97}
\end{figure}

Early on in its simulation the sideways flying saucer pancake shape
shows a clear breakdown of an unstable equilibrium oscillation.  We
have set the intial conditions so that MinD density is highly
concentrated to the left of a dividing line that runs parrallel to the
left vertical side.  For the first number of oscillations the maxima
trade off right to left, from a high peak in the rightward middle
compartment to vertically opposed peaks in the poles of the long axis
of the cell.  This is seen in the first part of Figure~\ref{box-97},
where the lower and upper sections peak in unison, and the middle
right section peaks $\pi$ radians removed from these.  After a few
oscillations the back and forth oscillations tip toward the vertical
direction and the system loses its initial oscillation pattern.  It
falls quickly instead into a back and forth pattern between the two
opposing endcaps.  The oscillation at this point is reminiscent of the
pills shape oscillation, once again showing the robust nature of the
back and forth, two pole pattern.

During each period, directly after one of the polar maxima has
subsided, there is a small spike with the regular maxima pattern (a
spike in membrane bound MinD:ATP immediately followed by a spike in
membrane bound MinD:MinE:ATP) seen in the small middle right
compartment section.  The animations show that the MinD, oscillating
back and forth between the upper and lower poles, seems to catch and
lag in the indentation for a bit before moving on.  Once again the
positive radius of curvature acts as an attractive place for MinD to
cluster on the walls, and it is not released to move on through the
cytoplasm again until MinE has formed its ring and completely reacted
it away.

\subsubsection{Randst-96-Star}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-700-700-9600-1500}
  %%\includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9600-1500}
  \caption{Total protein fluctuation in the star cell shape.  The vertical axis shows stacked the total
    number of proteins that are of four different compound stages and
    in the different sections of the cell.}
  \label{box-96}
\end{figure}

This clover shape finds a regular oscillatory behavoir after about
seven periods that consists of two maxima that peak regularly in the
corners accross from each other at the same times (Figure
\ref{box-96}) and then again opposing maxima that peak at the same
time in the other two corners.\fixme{should try some different
  starting conditions with this to see if there's another patter.}

Looking at at few periods in detail.  We find that there is a small
increase in cytoplasmic MinE always directly before the beggining
of climb in the membrane bound MinE peak.  Also, there is at any one
time a very large amount more of minE that is membrane bound then
cytoplasmic in any ones section.  A difference of a factor of roughly
20 when membrane MinE is at its peak, and roughly 12 when membrane
MinE is at it's minimum.  There seem to be two modes of oscillation,
each exhibiting a density peak maxima that moves back and forth
diagonaly from one corner to the opposite, as if there were two
overlapping, criss-crossed cells oscillating.  In the box plot this is
seen as roughly coinciding membrane minD peaks in the lower left and
right corners, followed one pi later in peaks in the upper right and
upper left corners.  Interesting that this did not exhibit
counterclockwise motion that Huangs 2008 paper might suggest.  Should
run sims that start in the wacky way that just started running with
triangles.
\newline

We haven't simulated long enough yet, so no real pattern has shown
itself.  There's certainly oscillations, and it's almost as if there's
competing modes - top to bottom, left to right, corner to corner, etc.
Interesting.

\fixme{maybe?}The animations seem to relax into a pattern in which the oscillations
go from the smallest corner, out to the other three, and then back to
the smallest corner?  \fixme{There isn't enough data to see properly
  though, need to run for longer.}

>>>>

\section{Stability Analysis and Mannik's Experimental Shapes}
In general we have found that the protein reaction sequence discussed
above results in extremely robust oscillatory behavoir of polar
selection and oscillation in a wide variety of cell shapes.  We have
also found that while all sizes of cell will in time settle into
regular oscillations, there are size dependent distinctions in the
number and placement of the spatial positions of maximization.
Regular oscillatory behavoir between two poles at opposite ends of the
longer cellular direction is found in cells which are long enough in
one dimension for the reaction chain to be unstable but short enough
in the perpendicular dimension to be stable.  The limits are seen when
both dimensions are too short to be unstable, so that initial
assymetry in protein concentration relaxes into a motionless steady
state (see Figure~\ref{box-mannik}, or when both are too great, so
that the proteins exhibit more than two poles and those at extreme
postitions in both dimensions.

In order to explain this theoretically, we have performed stability
analysis of the five differential equations in an infinite slab that
show a characteristic half wavelength in perturbative spatial
oscillations at which the system becomes unstable.  This length is
dependent upon slab thickness as shown in
Figure~\ref{fig:stability-analysis}.  When increasing the lengths and
widths of our simulated flattened cells, the cells stop exhibiting two
pole oscillalitory behavior and instead exhibit multiple
positions of maximization when the shortest distance across the cell
is approximately equal to the characteristic half wavelength at the
particular thickness at which we simulate (our simulations of cells
with thickness $0.25\mu m$ corresponds to half wavelengths of $2.13\mu
m$).  Huang~\cite{huang2003dynamic} performs a similar analysis for a
cylindrical shape \fixme{is it cylindrical or 1d?} and finds
characteristic half wavelengths of $2\mu m$.

\begin{figure}
  \includegraphics[width=4.2cm]{../data/shape-randst/plots/box-plot_D--randst-25-800-800-9500-1500}
  \includegraphics[width=4.2cm]{../data/shape-randst/plots/box-plot_D--randst-25-1000-1700-9400-1500}
  \caption{Box plots of Mannik shapes for sizes at the lower stability
    threshhold.  The vertical axis shows stacked and changing over
    time the total number of proteins that are of four different
    compound stages and in the different sections of the cell.  Both
    shapes have longer axis of roughly 2.2$\mu m$.}
  \label{box-mannik}
\end{figure}

Decreasing the cell size in succesive steps, we can see the critical
size at which the cell oscillations are quickly damped out.
Figure~\ref{box-mannik} shows box plots of the largest cells simulated
whose oscillations relax into the steady state solution.  Both have a
long axis of roughly 2.2$\mu m$, in agreement with theory.

\begin{figure}
  \includegraphics[width=2.8cm]{../mannik-2.png}
  \includegraphics[width=2.8cm]{../mannik-1.png}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/paper-arrow-plot}
  \caption{Arrow plots for our simulation and mannik's experimental}
  \label{arrow-compare-mannik}
\end{figure}

Figure~\ref{arrow-compare-mannik} shows plots of total MinD
concentration maxima that are global in space and local in time.  The
tip of an arrow represents one of these maxima and the adjacent arrow
(whose tail is touching the previous arrow's tip) points to the next
such maxima in time.  Mannik \emph{et al.} have published similar
plots of two of their cells, and we have attempted to replicate the
shape of these cells in our simulation in order to compare.  We find
that when we replicate the size as well as shape of their cells as
near as possible, we do not see the disordered protein behavoir that
they observe.  Instead we see a robust ability of the cell to find two
prefered pole locations and to oscillate maxima back and forth very
steadily between them.  When we increase the size of our cells, we
indeed see multiple positions of maxima develop, but we never observe
the random behavoir that they have observed experimentally.  This is
perhaps due to the exact, continuous method we use when conducting
simulations.  Stochastic methods would perhaps show more random
behavoir at these sizes.  We consider this to be a limitation of the
exact solution method.

@@

\begin{figure}
  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-233-233-233-1500}
  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-266-266-266-1500}
  \includegraphics[width=4.2cm]{../data/shape-triangle/plots/box-plot_D--triangle-25-800-800-800-1500}
  \caption{Two box plots of triangle shapes.  The two smaller show
    sizes at the lower stability threshhold, one just below, one just
    above.  The larger triangle is the largest simulated. The vertical
    axis shows stacked and changing over time the total number of
    proteins that are of four different compound stages and in the
    different sections of the cell.}
  \label{box-triangle}
\end{figure}

Along with our other shapes we have chosen to simulate both
equilateral and iscoscolese triangular pancake shapes, since these
offer in a way the simplest shapes that are geometrically more
complicated than our pill shapes.

Both triangle types show a similar oscillation pattern.  The
iscosolese exhibit regular oscillations characterized by one distinct
pole in the more accute angle of the three, and another 'pole' along
the opposite wall, peaking in the two more obtuse angles at the same
time.  The equilateral triangles settle very quickly into the same
pattern, where the choice of which is one pole corner is dependent
upon initial conditions.  As with the Mannik simulated shapes, there
is a critical small size below which the oscillations dampen out and
will not achieve a regular pattern.  Figure~\ref{box-triangle} shows
three plots of equilateral triangles.  The two smaller show the
largest that showed damped out behavoir and the smallest that
developed regular oscillations.  The larger shows the largest traingle
that we simulated, showing that at even very large sizes the behavoir
is consistent and stable.  Above the small size threshold, all
simulated triangles show regular patterns for as long as simulated
(roughly 800 seconds).

@@

\section{Our 'V' Shaped Cells}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-500-500-9900-1500}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-525-525-9900-1500}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-550-550-9900-1500}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9900-1500}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-1400-1400-9900-1500}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-1500-1500-9900-1500}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-1600-1600-9900-1500}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-1800-1800-9900-1500}
  \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-2000-2000-9900-1500}
\end{figure}

@@

%% \subsubsection{Randst-97-}
%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-800-600-9700-1500}
%%   %%\includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9600-1500}
%%   \caption{Total protein fluctuation in the sideways flying saucer
%%     pancake cell shape.  The vertical axis shows stacked the total
%%     number of proteins that are of four different compound stages and
%%     in the different sections of the cell.}
%%   \label{box-97}
%% \end{figure}

%% Early on in its simulation the sideways flying saucer pancake shape
%% shows a clear breakdown of an unstable equilibrium oscillation.  We
%% have set the intial conditions so that MinD density is highly
%% concentrated to the left of a dividing line that runs parrallel to the
%% left vertical side.  For the first number of oscillations the maxima
%% trade off right to left, from a high peak in the rightward middle
%% compartment to vertically opposed peaks in the poles of the long axis
%% of the cell.  This is seen in the first part of Figure~\ref{box-97},
%% where the lower and upper sections peak in unison, and the middle
%% right section peaks $\pi$ radians removed from these.  After a few
%% oscillations the back and forth oscillations tip toward the vertical
%% direction and the system loses its initial oscillation pattern.  It
%% falls quickly instead into a back and forth pattern between the two
%% opposing endcaps.  The oscillation at this point is reminiscent of the
%% pills shape oscillation, once again showing the robust nature of the
%% back and forth, two pole pattern.

%% During each period, directly after one of the polar maxima has
%% subsided, there is a small spike with the regular maxima pattern (a
%% spike in membrane bound MinD:ATP immediately followed by a spike in
%% membrane bound MinD:MinE:ATP) seen in the small middle right
%% compartment section.  The animations show that the MinD, oscillating
%% back and forth between the upper and lower poles, seems to catch and
%% lag in the indentation for a bit before moving on.  Once again the
%% positive radius of curvature acts as an attractive place for MinD to
%% cluster on the walls, and it is not released to move on through the
%% cytoplasm again until MinE has formed its ring and completely reacted
%% it away.

%% \subsubsection{Randst-96-Star}
%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-700-700-9600-1500}
%%   %%\includegraphics[width=\columnwidth]{../data/shape-randst/plots/box-plot_D--randst-25-600-600-9600-1500}
%%   \caption{Total protein fluctuation in the star cell shape.  The vertical axis shows stacked the total
%%     number of proteins that are of four different compound stages and
%%     in the different sections of the cell.}
%%   \label{box-96}
%% \end{figure}

%% This clover shape finds a regular oscillatory behavoir after about
%% seven periods that consists of two maxima that peak regularly in the
%% corners accross from each other at the same times (Figure
%% \ref{box-96}) and then again opposing maxima that peak at the same
%% time in the other two corners.\fixme{should try some different
%%   starting conditions with this to see if there's another patter.}

%% Looking at at few periods in detail.  We find that there is a small
%% increase in cytoplasmic MinE always directly before the beggining
%% of climb in the membrane bound MinE peak.  Also, there is at any one
%% time a very large amount more of minE that is membrane bound then
%% cytoplasmic in any ones section.  A difference of a factor of roughly
%% 20 when membrane MinE is at its peak, and roughly 12 when membrane
%% MinE is at it's minimum.  There seem to be two modes of oscillation,
%% each exhibiting a density peak maxima that moves back and forth
%% diagonaly from one corner to the opposite, as if there were two
%% overlapping, criss-crossed cells oscillating.  In the box plot this is
%% seen as roughly coinciding membrane minD peaks in the lower left and
%% right corners, followed one pi later in peaks in the upper right and
%% upper left corners.  Interesting that this did not exhibit
%% counterclockwise motion that Huangs 2008 paper might suggest.  Should
%% run sims that start in the wacky way that just started running with
%% triangles.
%% \newline

%% We haven't simulated long enough yet, so no real pattern has shown
%% itself.  There's certainly oscillations, and it's almost as if there's
%% competing modes - top to bottom, left to right, corner to corner, etc.
%% Interesting.

%% \fixme{maybe?}The animations seem to relax into a pattern in which the oscillations
%% go from the smallest corner, out to the other three, and then back to
%% the smallest corner?  \fixme{There isn't enough data to see properly
%%   though, need to run for longer.}
